---
title: "A final analysis of our HapMap data"
description: |
    How we combined our knowledge of multiple hypothesis testing and PLINK results to determine an appropriate significance threshold for our HapMap dataset
output: 
  distill::distill_article:
    toc: true
bibliography: Library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Learn more about creating websites with Distill at:
# https://rstudio.github.io/distill/website.html
```

```{r}
library(tidyverse)
library(janitor)
library(snpStats)
library(NatParksPalettes)
```

https://www.bioconductor.org/packages/devel/bioc/vignettes/SNPRelate/inst/doc/SNPRelate.html
http://sites.tufts.edu/cbi/files/2013/02/GWAS_Exercise6_Stratification.pdf

# Hapmap data

## Data loading 

```{r}
fam <- 'hapmapData/HapMap_3_r3_1.fam'
bim <- 'hapmapData/HapMap_3_r3_1.bim'
bed <- 'hapmapData/HapMap_3_r3_1.bed'

hapmap <- read.plink(bed, bim, fam)
```

## Create manhattan plot

Create a null trait and write a file to the plink data folder.

```{r}
hapmap_nulltrait <- cbind(hapmap$fam, trait = rnorm(n = 165, mean = 0, sd = 1)) %>%
  dplyr::select(pedigree, member, trait)
write_delim(hapmap_nulltrait, "hapmapData/hapmap_nulltrait")
```

In PLINK run ./plink --bfile HapMap_3_r3_1 --assoc --adjust --pheno hapmap_nulltrait --out as2

Read in results from PLINK:

```{r}
hapmap_adjusted <- read_table("hapmapData/as2.qassoc.adjusted")
hapmap_unadjusted <- read_table("hapmapData/as2.qassoc")

results_with_position_adjusted <- hapmap_adjusted %>%
  mutate(CHR = as.integer(CHR)) %>%
  left_join(hapmap$map %>%
              dplyr::select(snp.name, position, chromosome), by = c("SNP" = "snp.name", "CHR" = "chromosome"))

results_with_position_unadjusted <- hapmap_unadjusted %>%
  mutate(CHR = as.integer(CHR)) %>%
  left_join(hapmap$map %>%
              dplyr::select(snp.name, position, chromosome), by = c("SNP" = "snp.name", "CHR" = "chromosome"))
```

Create manhattan plot for adjusted and unadjusted results

```{r, eval=FALSE}
results_with_position_adjusted %>%
  mutate(minuslogp = -log10(GC),
         CHR = as.factor(CHR)) %>%
  ggplot(aes(x = CHR, y = minuslogp, group = interaction(CHR, position), color = CHR)) + 
  geom_point(position = position_dodge(0.8)) + 
  labs(x = 'chromosome', y = expression(paste('-log'[10],'(p-value)')))+
  theme_classic()+
  geom_hline(yintercept = 7.081506, color = "navy", linetype = "dashed")+
  geom_hline(yintercept = 1.30103, color = "navy", linetype = "dashed")+
  scale_color_manual(values=natparks.pals("DeathValley",24))+
  theme(legend.position = "none")
```

```{r, eval=FALSE}
results_with_position_unadjusted %>%
  mutate(minuslogp = -log10(GC),
         CHR = as.factor(CHR)) %>%
  ggplot(aes(x = CHR, y = minuslogp, group = interaction(CHR, position), color = CHR)) + 
  geom_point(position = position_dodge(0.8)) + 
  labs(x = 'chromosome', y = expression(paste('-log'[10],'(p-value)')))+
  theme_classic()+
  geom_hline(yintercept = 7.081506, color = "navy", linetype = "dashed")+
  geom_hline(yintercept = 1.30103, color = "navy", linetype = "dashed")+
  scale_color_manual(values=natparks.pals("DeathValley",24))+
  theme(legend.position = "none")
```

## Complete replications

Determine a threshold using simulation. Create a function to generate a quantative trait 165 times. Next, replicate that trait 1000 times and bind it the memberId and pedigree information. Write this file to folder to use in plink

```{r, eval=FALSE}
create_quantitative_trait <- function(i){
  y <- rnorm(n = 165, mean = 0, sd = 1) 
}

traits <- as.data.frame(replicate(1000, create_quantitative_trait()))

hapmap_traits <- cbind(hapmap$fam %>%
        dplyr::select(1:2), traits)

write_delim(hapmap_traits, "hapmapData/hapmap_traits")
```

In plink, run ./plink --bfile HapMap_3_r3_1 --assoc --pheno hapmap_traits --all-pheno --pfilter 1e-3

Running that PLINK command takes about 15 minutes. Load the data files back into RStudio (takes about 3 minutes). Take the minimum p-value in each file and plot the distribution. Take the 5% quantile of the minimum p-values: this is our threshold.

```{r, cache=TRUE}
dataFiles <- lapply(Sys.glob("hapmapData/plink.P*.qassoc"), read_table)

pvalues <- sapply(dataFiles, function(x) min(x$P, na.rm=TRUE))

as.data.frame(pvalues) %>%
  ggplot(aes(x=pvalues))+
  geom_density(fill = "cadetblue")+
  theme_classic()+
  annotate(geom = "text", color = "red", x = 1e-05, y = 250000, label = "0.05 quantile:\n7.31495e-08", family = "mono", cex = 3)+
  geom_vline(xintercept = 7.31495e-08, color = "red", linetype = "dashed")+
  labs(x="P-values", y = "Density", title = "Distribution of minimum p-values for 1000 replications")+
  geom_curve(aes(x = 1e-05, y = 270000, xend = 1e-06, yend = 350000), 
             arrow = arrow(length = unit(0.03, "npc")), curvature = 0.3, color = "red")+
  theme(plot.title.position = "plot", 
        plot.title = element_text(family = "mono"), 
        axis.title = element_text(family = "mono"), 
        axis.text = element_text(family = "mono"))

quantile(pvalues, 0.05)
```

We get a threshold of 8.28885e-08 for the Hapmap data.

# R-bloggers Data

In PLINK, run plink --merge-list allfiles.txt --make-bed --out rbloggersComb

```{r}
load("rbloggersData/conversionTable.RData")

pathM <- paste("rbloggersData/108Malay_2527458snps", c(".bed", ".bim", ".fam"), sep = "")
SNP_M <- read.plink(pathM[1], pathM[2], pathM[3])

pathI <- paste("rbloggersData/105Indian_2527458snps", c(".bed", ".bim", ".fam"), sep = "")
SNP_I <- read.plink(pathI[1], pathI[2], pathI[3])

pathC <- paste("rbloggersData/110Chinese_2527458snps", c(".bed", ".bim", ".fam"), sep = "")
SNP_C <- read.plink(pathC[1], pathC[2], pathC[3])

rbloggers_fam <- rbind(SNP_M$fam, SNP_I$fam, SNP_C$fam)
rbloggers_map <- rbind(SNP_M$map, SNP_I$map, SNP_C$map)
```

```{r}
rbloggers_nulltrait <- cbind(rbloggers_fam %>% dplyr::select(1:2), trait = rnorm(n = 323, mean = 0, sd = 1))
write_delim(rbloggers_nulltrait, "rbloggersData/rbloggers_nulltrait")
```

In PLINK run ./plink --bfile rbloggersComb --assoc --adjust --pheno rbloggers_nulltrait --out as2

Read in results from PLINK:

```{r}
rbloggers_adjusted <- read_table("rbloggersData/as2.qassoc.adjusted")
rbloggers_unadjusted <- read_table("rbloggersData/as2.qassoc")

rbloggers_adjusted_position <- rbloggers_adjusted%>%
  mutate(CHR = as.integer(CHR)) %>%
  left_join(rbloggers_map %>%
              dplyr::select(snp.name, position, chromosome), by = c("SNP" = "snp.name", "CHR" = "chromosome"))

rbloggers_unadjusted_position <- rbloggers_unadjusted%>%
  mutate(CHR = as.integer(CHR)) %>%
  left_join(rbloggers_map %>%
              dplyr::select(snp.name, position, chromosome), by = c("SNP" = "snp.name", "CHR" = "chromosome"))
```

Create manhattan plots:

```{r, cache=TRUE}
rbloggers_adjusted_position %>%
  mutate(minuslogp = -log10(GC),
         CHR = as.factor(CHR)) %>%
  ggplot(aes(x = CHR, y = minuslogp, group = interaction(CHR, position), color = CHR)) + 
  geom_point(position = position_dodge(0.8)) + 
  labs(x = 'chromosome', y = expression(paste('-log'[10],'(p-value)')))+
  theme_classic()+
  geom_hline(yintercept = 7.26566, color = "navy", linetype = "dashed")+
  geom_hline(yintercept = 1.30103, color = "navy", linetype = "dashed")+
  scale_color_manual(values=natparks.pals("DeathValley",24))+
  theme(legend.position = "none")
```

## Complete replications

```{r, eval=FALSE}
create_quantitative_trait <- function(i){
  y <- rnorm(n = 323, mean = 0, sd = 1) 
}

traits <- as.data.frame(replicate(1000, create_quantitative_trait()))

rbloggers_traits <- cbind(rbloggers_fam %>%
        dplyr::select(1:2), traits)

write_delim(rbloggers_traits, "rbloggersData/rbloggers_traits")
```

./plink --bfile rbloggersComb --assoc --pheno rbloggers_traits --all-pheno --pfilter 1e-3

```{r, cache=TRUE}
dataFiles <- lapply(Sys.glob("rbloggersData/plink.P*.qassoc"), read_table)

pvalues <- sapply(dataFiles, function(x) min(x$P, na.rm=TRUE))

as.data.frame(pvalues) %>%
  ggplot(aes(x=pvalues))+
  geom_density(fill = "cadetblue")+
  theme_classic()+
  annotate(geom = "text", color = "red", x = 1e-05, y = 250000, label = "0.05 quantile:\n7.31495e-08", family = "mono", cex = 3)+
  geom_vline(xintercept = 7.31495e-08, color = "red", linetype = "dashed")+
  labs(x="P-values", y = "Density", title = "Distribution of minimum p-values for 1000 replications")+
  geom_curve(aes(x = 1e-05, y = 270000, xend = 1e-06, yend = 350000), 
             arrow = arrow(length = unit(0.03, "npc")), curvature = 0.3, color = "red")+
  theme(plot.title.position = "plot", 
        plot.title = element_text(family = "mono"), 
        axis.title = element_text(family = "mono"), 
        axis.text = element_text(family = "mono"))

quantile(pvalues, 0.05)
```

5.42425e-08

# Finally, R-bloggers data with varying populations

Create a trait varying between the populations

```{r}
rbloggers_poptrait <- cbind(rbloggers_fam %>% select(1:2), trait = c(rnorm(n = 108, mean = 0, sd = 1), rnorm(105, -1, sd = 1), rnorm(110, 1, 1)))
write_delim(rbloggers_poptrait, "rbloggersSep/rbloggers_poptrait")
```

In PLINK run ./plink --bfile rbloggersComb --assoc --adjust --pheno rbloggers_poptrait --out as2

```{r}
rbloggersSep_adjusted <- read_table("rbloggersSep/as2.qassoc.adjusted")
rbloggersSep_unadjusted <- read_table("rbloggersSep/as2.qassoc")

rbloggersSep_adjusted_position <- rbloggersSep_adjusted%>%
  mutate(CHR = as.integer(CHR)) %>%
  left_join(rbloggers_map %>%
              dplyr::select(snp.name, position, chromosome), by = c("SNP" = "snp.name", "CHR" = "chromosome"))

rbloggersSep_unadjusted_position <- rbloggersSep_unadjusted%>%
  mutate(CHR = as.integer(CHR)) %>%
  left_join(rbloggers_map %>%
              dplyr::select(snp.name, position, chromosome), by = c("SNP" = "snp.name", "CHR" = "chromosome"))
```

```{r, cache=TRUE}
rbloggersSep_adjusted_position %>%
  mutate(minuslogp = -log10(GC),
         CHR = as.factor(CHR)) %>%
  ggplot(aes(x = CHR, y = minuslogp, group = interaction(CHR, position), color = CHR)) + 
  geom_point(position = position_dodge(0.8)) + 
  labs(x = 'chromosome', y = expression(paste('-log'[10],'(p-value)')))+
  theme_classic()+
  geom_hline(yintercept = 7.245961, color = "navy", linetype = "dashed")+
  geom_hline(yintercept = 1.30103, color = "navy", linetype = "dashed")+
  scale_color_manual(values=natparks.pals("DeathValley",24))+
  theme(legend.position = "none")
```

## Complete replications

```{r, eval=FALSE}
create_quantitative_trait <- function(i){
  c(rnorm(n = 108, mean = 0, sd = 1), rnorm(105, -1, sd = 1), rnorm(110, 1, 1)) 
}

traits <- as.data.frame(replicate(1000, create_quantitative_trait()))

rbloggers_Poptraits <- cbind(rbloggers_fam %>%
        dplyr::select(1:2), traits)

write_delim(rbloggers_Poptraits, "rbloggersSep/rbloggers_Poptraits")
```

./plink --bfile rbloggersComb --assoc --pheno rbloggers_Poptraits --all-pheno --pfilter 1e-3

Read files back into R

```{r, cache=TRUE}
dataFiles <- lapply(Sys.glob("rbloggersSep/plink.P*.qassoc"), read_table)

pvalues <- sapply(dataFiles, function(x) min(x$P, na.rm=TRUE))

as.data.frame(pvalues) %>%
  ggplot(aes(x=pvalues))+
  geom_density(fill = "cadetblue")+
  theme_classic()+
  annotate(geom = "text", color = "red", x = 1e-05, y = 250000, label = "0.05 quantile:\n7.31495e-08", family = "mono", cex = 3)+
  geom_vline(xintercept = 7.31495e-08, color = "red", linetype = "dashed")+
  labs(x="P-values", y = "Density", title = "Distribution of minimum p-values for 1000 replications")+
  geom_curve(aes(x = 1e-05, y = 270000, xend = 1e-06, yend = 350000), 
             arrow = arrow(length = unit(0.03, "npc")), curvature = 0.3, color = "red")+
  theme(plot.title.position = "plot", 
        plot.title = element_text(family = "mono"), 
        axis.title = element_text(family = "mono"), 
        axis.text = element_text(family = "mono"))

quantile(pvalues, 0.05)
```
